---
title: "microstegium_model_kortessis_comparison"
author: "Amy Kendig"
date: "11/20/2019"
output: pdf_document
---

```{r options, include = F}
library(knitr)
# wrap text
opts_chunk$set(tidy.opts=list(width.cutoff = 70), tidy = T)
```

## Summary
The goal is to compare my model with Nick's, which are the same except that:  
- I model decomposition as exponential decay between years instead of proportion leftover (1-d) - changed mine to match  
- I made litter suppression linear with a manual cut-off instead of using the Beverton-Holt form  - changed mine to match  
- I use an establishment rate (i.e., summer survival) instead of a variable representing size (biomass/seed) for both annual and perennial. I don't think the relationship between size and seed production changes with competition. These variables are both 1 below (i.e., no scenario testing).  
- I allow for different competitive effects of different species. I set them equal to each other for this (i.e., no scenario testing).  
- My transition from seed count to biomass production is different because I convert the individuals to biomass and Nick's model has biomass as a component of the annual seed model (vigor). He also uses an overwinter decomposition rate (my litter build-up is slower because the conversion parameter is so small).  
    - Scenario 3 (aklitter): large change in density  
- I include establishment (for seeds) and winter survival (for adults) in the competition pressure.  
    - Scenario 4 (akcomp): minor change in density
- The perennial model is different because I have first-year seedlings producing seeds and experiencing competition. They have the option to cause competitive effects, but I set that to 0.  
    - Scenario 5 (akper): no change in density  
- I account for perennial adult survival before calculating seed production.  
    - Scenario 6 (akseed): large change in density.  
- A change to make to both models? The per-seed contribution to litter should change with competition (and potentially disease).  
- A change to make to both models: water loss when coverting to seeds to biomass.  
- Future direction: competition between pathogen strains on the plants, coexistence of different pathogen strains  


## Set-up
```{r set-up, message = F, warning = F}
# clear all existing data
rm(list=ls())

# open libraries
library(data.table)
library(plotly)
library(cowplot)
library(popbio)
library(tidyverse)

# plotting parameters
axisText=10
axisTitle=12
legendText=10
legendTitle=0

# import parameters
params <- read_csv("../data/mv_ev_kortessis_comparison_111919.csv", col_types = cols())
```

## Parameters

#### Formatting:
* subscripts follow "."
* p = perennial adult (at least 1 year old)
* s = perennial seedling (germinated that spring)
* a = annual
* L = annual litter

#### Settings:
```{r settings, tidy = F}
# initial conditions
N0.a = 5
N0.s = 5
N0.p = 5
L0 = 2

# simulation time
years = 300
```

```{r parameters-table, echo = F, results = 'asis'}
# other parameter values
kable(params, caption = "Model parameter values")
```

## Amy's model description

#### Population equations:
N.a[t+1] = s.a * (1-g.a[t]) * N.a[t] + g.a[t] * h.a * f.a[t] * N.a[t]  
annual seeds = seed bank survival + seed production  
  
L[t+1] = c.a * g.a[t] * h.a[t] * N.a[t] + L[t] * (1 - b) 
annual litter = biomass from previous fall + decomposition   

Assume counts are being conducted in the fall  
N.s[t+1] = s.s * (1-g.s[t]) * N.s[t] + g.s[t] * h.s * f.s[t] * N.s[t] + m.p * f.p[t] * N.p[t]  
perennial seeds = seed bank survival + seedling seed production + adult seed production  
  
N.p[t+1] = m.p * N.p[t] + g.s[t] * h. s * N.s[t]  
perennial adults = survival + seedling maturation  
  
### Density-dependence on fecundity:
f.s[t] = lambda.s / (1 + alpha.ss * g.s[t] * h.s * N.s[t] + alpha.sp * m.p * N.p[t] + alpha.sa * g.a[t] * h.a * N.a[t])  
perennial seedling fecundity = fecundity in the absence of competition / (perennial seedling competition + perennial adult competition + annual competition)

f.p[t] = lambda.p / (1 + alpha.ps * g.s[t] * h.s * N.s[t] + alpha.pp * m.p * N.p[t] + alpha.pa * g.a[t] * h.a * N.a[t])  
perennial adult fecundity = fecundity in the absence of competition / (perennial seedling competition + perennial adult competition + annual competition)

f.a[t] = lambda.a / (1 + alpha.as * g.s[t] * h.s * N.s[t] + alpha.ap * m.p * N.p[t] + alpha.aa * g.a[t] * h.a * N.a[t])  
annual fecundity = fecundity in the absence of competition / (perennial seedling competition + perennial adult competition + annual competition)

#### Litter suppression:
g.a[t] = gamma.a / (1 + alpha.aL * L[t])  
annual seed germination = germination in the absence of litter / reduction due to litter    

g.s[t] = gamma.s / (1 + alpha.sL * L[t])  
perennial seed germination = germination in the absence of litter / reduction due to litter  

#### Amy's model function
```{r amy-model}
simFun_a = function(params, N0.a, N0.s, N0.p, L0, simtime, alt.comp, alt.per, alt.seed){
  
  # define parameters
  m.p = filter(params, symbol == "m.p")$value
  s.a = filter(params, symbol == "s.a")$value
  s.s = filter(params, symbol == "s.s")$value
  gamma.a = filter(params, symbol == "gamma.a")$value
  gamma.s = filter(params, symbol == "gamma.s")$value
  alpha.aL = filter(params, symbol == "alpha.aL")$value
  alpha.sL = filter(params, symbol == "alpha.sL")$value
  b = filter(params, symbol == "b")$value
  h.a = filter(params, symbol == "h.a")$value
  h.s = filter(params, symbol == "h.s")$value
  h.p = filter(params, symbol == "h.p")$value
  v.a = filter(params, symbol == "v.a")$value
  v.p = filter(params, symbol == "v.p")$value
  h.p = filter(params, symbol == "h.p")$value
  lambda.a = filter(params, symbol == "lambda.a")$value
  lambda.p = filter(params, symbol == "lambda.p")$value
  lambda.s = filter(params, symbol == "lambda.s")$value
  alpha.aa = filter(params, symbol == "alpha.aa")$value
  alpha.as = filter(params, symbol == "alpha.as")$value
  alpha.ap = filter(params, symbol == "alpha.ap")$value
  alpha.sa = filter(params, symbol == "alpha.sa")$value
  alpha.ss = filter(params, symbol == "alpha.ss")$value
  alpha.sp = filter(params, symbol == "alpha.sp")$value
  alpha.pa = filter(params, symbol == "alpha.pa")$value
  alpha.ps = filter(params, symbol == "alpha.ps")$value
  alpha.pp = filter(params, symbol == "alpha.pp")$value
  c.a = filter(params, symbol == "c.a")$value
  w.a = filter(params, symbol == "w.a")$value
  
  # initialize populations
  N.a = rep(NA,simtime)
  N.s = rep(NA,simtime)
  N.p = rep(NA,simtime)
  L = rep(NA,simtime)
  
  N.a[1] = N0.a
  N.s[1] = N0.s
  N.p[1] = N0.p
  L[1] = L0
  
  # simulate population dynamics
  for(t in 1:(simtime-1)){	
    
    # reduce germination due to litter
    g.a = gamma.a/(1+alpha.aL*L[t])
    g.s = gamma.s/(1+alpha.sL*L[t])
    
    # reduce fecundity due to competition (include survival)
    f.a1 = lambda.a/(1+alpha.as*g.s*h.s*N.s[t]+alpha.ap*m.p*N.p[t]+alpha.aa*g.a*h.a*N.a[t])
    f.s1 = lambda.s/(1+alpha.ss*g.s*h.s*N.s[t]+alpha.sp*m.p*N.p[t]+alpha.sa*g.a*h.a*N.a[t])
    f.p1 = lambda.p/(1+alpha.ps*g.s*h.s*N.s[t]+alpha.pp*m.p*N.p[t]+alpha.pa*g.a*h.a*N.a[t])
    
    # reduce fecundity due to competition (don't include survival)
    f.a2 = lambda.a/(1+alpha.as*g.s*N.s[t]+alpha.ap*N.p[t]+alpha.aa*g.a*N.a[t])
    f.s2 = lambda.s/(1+alpha.ss*g.s*N.s[t]+alpha.sp*N.p[t]+alpha.sa*g.a*N.a[t])
    f.p2 = lambda.p/(1+alpha.ps*g.s*N.s[t]+alpha.pp*N.p[t]+alpha.pa*g.a*N.a[t])
    
    # choose fecundity 
    f.a = ifelse(alt.comp == T, f.a2, f.a1)
    f.s = ifelse(alt.comp == T, f.s2, f.s1)
    f.p = ifelse(alt.comp == T, f.p2, f.p1)
    
    # remove first-year seed production
    f.s = ifelse(alt.per == T, 0, f.s)
    
    # choose adult seed production
    seed.p = ifelse(alt.seed == T, f.p, m.p*f.p)

    # population size
    N.a[t+1] = s.a*(1-g.a)*N.a[t]+g.a*h.a*f.a*N.a[t]
    L[t+1] = c.a*g.a*h.a*N.a[t]+(1-b)*L[t]
    N.s[t+1] = s.s*(1-g.s)*N.s[t]+g.s*h.s*f.s*N.s[t]+seed.p*N.p[t]
    N.p[t+1] = m.p*N.p[t]+g.s*h.s*N.s[t]
    
    # correct to prevent negative numbers
    N.a[t+1] = ifelse(N.a[t+1]<1,0,N.a[t+1])
    L[t+1] = ifelse(L[t+1]<0,0,L[t+1])
    N.s[t+1] = ifelse(N.s[t+1]<1,0,N.s[t+1])
    N.p[t+1] = ifelse(N.p[t+1]<1,0,N.p[t+1])
  }
  
  # total perennial population
  N.t = N.s+N.p
  
  # save data
  dfN = data.frame(time = rep(1:simtime,5),
                   N = c(N.a, L, N.s, N.p, N.t),
                   species = rep(c("Annual", "Annual litter","Perennial seedling", "Perennial adult",  "Perennial"), each = simtime))
  
  # return
  return(dfN)
}
```

#### Nick's model function
```{r nick-model}
simFun_n = function(params, N0.a, N0.s, N0.p, L0, simtime){
  
  # define parameters
  m.p = filter(params, symbol == "m.p")$value
  s.a = filter(params, symbol == "s.a")$value
  s.s = filter(params, symbol == "s.s")$value
  gamma.a = filter(params, symbol == "gamma.a")$value
  gamma.s = filter(params, symbol == "gamma.s")$value
  alpha.aL = filter(params, symbol == "alpha.aL")$value
  alpha.sL = filter(params, symbol == "alpha.sL")$value
  b = filter(params, symbol == "b")$value
  h.a = filter(params, symbol == "h.a")$value
  h.s = filter(params, symbol == "h.s")$value
  h.p = filter(params, symbol == "h.p")$value
  v.a = filter(params, symbol == "v.a")$value
  v.p = filter(params, symbol == "v.p")$value
  lambda.a = filter(params, symbol == "lambda.a")$value
  lambda.p = filter(params, symbol == "lambda.p")$value
  lambda.s = filter(params, symbol == "lambda.s")$value
  alpha.aa = filter(params, symbol == "alpha.aa")$value
  alpha.as = filter(params, symbol == "alpha.as")$value
  alpha.ap = filter(params, symbol == "alpha.ap")$value
  alpha.sa = filter(params, symbol == "alpha.sa")$value
  alpha.ss = filter(params, symbol == "alpha.ss")$value
  alpha.sp = filter(params, symbol == "alpha.sp")$value
  alpha.pa = filter(params, symbol == "alpha.pa")$value
  alpha.ps = filter(params, symbol == "alpha.ps")$value
  alpha.pp = filter(params, symbol == "alpha.pp")$value
  c.a = filter(params, symbol == "c.a")$value
  w.a = filter(params, symbol == "w.a")$value
  
  # initialize populations
  N.a = rep(NA,simtime)
  N.s = rep(NA,simtime)
  N.p = rep(NA,simtime)
  L = rep(NA,simtime)
  
  N.a[1] = N0.a
  N.s[1] = N0.s
  N.p[1] = N0.p
  L[1] = L0
  
  # simulate population dynamics
  for(t in 1:(simtime-1)){	
    
    # reduce germination due to litter
    g.a = gamma.a/(1+alpha.aL*L[t])
    g.s = gamma.s/(1+alpha.sL*L[t])
    
    # reduce fecundity due to competition
    f.a = lambda.a/(1+alpha.ap*N.p[t]+alpha.aa*g.a*N.a[t])
    f.p = lambda.p/(1+alpha.pp*N.p[t]+alpha.pa*g.a*N.a[t])

    # population size
    N.a[t+1] = s.a*(1-g.a)*N.a[t]+g.a*v.a*f.a*N.a[t]
    L[t+1] = w.a*(g.a*v.a*N.a[t]+(1-b)*L[t])
    N.s[t+1] = s.s*(1-g.s)*N.s[t]+v.p*f.p*N.p[t]
    N.p[t+1] = m.p*N.p[t]+g.s*N.s[t]
    
    # correct to prevent negative numbers
    N.a[t+1] = ifelse(N.a[t+1]<1,0,N.a[t+1])
    L[t+1] = ifelse(L[t+1]<0,0,L[t+1])
    N.s[t+1] = ifelse(N.s[t+1]<1,0,N.s[t+1])
    N.p[t+1] = ifelse(N.p[t+1]<1,0,N.p[t+1])
  }
  
  # total perennial population
  N.t = N.s+N.p
  
  # save data
  dfN = data.frame(time = rep(1:simtime,5),
                   N = c(N.a, L, N.s, N.p, N.t),
                   species = rep(c("Annual", "Annual litter","Perennial seedling", "Perennial adult",  "Perennial"), each = simtime))
  
  # return
  return(dfN)
}
```

## Simulations

#### Function for creating figures:
```{r figure_function, tidy = F}
figfun <- function(moda, modn, filename){
  
  # Format data
  mod <- moda %>%
  mutate(model = "Amy's") %>%
  full_join(modn %>% mutate(model = "Nick's"), by = c("time", "N", "species", "model")) %>%
  mutate(life_history = case_when(species %in% c("Annual", "Annual litter") ~ "annual",
         TRUE ~ "perennial"),
         status = case_when(species %in% c("Annual", "Perennial", "Perennial adult") ~ "annual/adults",
         TRUE ~ "litter/seedlings"),
         model = factor(model, levels = c("Nick's", "Amy's")))
  
  # Total population figure
  tot_fig <- mod %>%
  filter(species %in% c("Annual", "Perennial")) %>%
  ggplot(aes(x = time, y = N, color = life_history)) +
  geom_line(size = 1.5) + 
  facet_wrap(~model) +
  xlab("Time") +
  ylab("Density") +
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle),
        strip.text = element_text(size = axisTitle),
        strip.background = element_blank())
  
  # Each population figure
  each_fig <- mod %>%
  filter(species != "Perennial") %>%
  ggplot(aes(x = time, y = N, group = species, color = life_history, linetype = status)) +
  geom_line(size = 1.5) + 
  facet_wrap(~model) +
    xlab("Time") +
  ylab("Density") +
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle),
        strip.text = element_text(size = axisTitle),
        strip.background = element_blank())

  # Output figure
  print(plot_grid(tot_fig, each_fig, nrow = 2))
  
  # Save output figure
  pdf(filename)
  print(plot_grid(tot_fig, each_fig, nrow = 2))
  dev.off()
  
  # Output dataframe
  return(mod)
}

```

#### Original models (parameterization 1)
```{r nick-sim, tidy = F}
nmod <- simFun_n(params = params, 
       N0.a = N0.a, 
       N0.s = N0.s, 
       N0.p = N0.p, 
       L0 = L0, 
       simtime = years)
```

```{r amy-sim, tidy = F}
amod <- simFun_a(params = params, 
       N0.a = N0.a, 
       N0.s = N0.s, 
       N0.p = N0.p, 
       L0 = L0, 
       simtime = years,
       alt.comp = F,
       alt.per = F,
       alt.seed = F)
```

```{r orig, tidy = F}
mod1 <- figfun(amod, nmod, "./microstegium_model_kortessis_comparison_output_orig.pdf")
```

#### Make models equivalent (parameterization 2)
```{r amy-sim-2, tidy = F}
params2 <- params %>%
  mutate(value = case_when(symbol == "c.a" ~ 0.5,
                           symbol == "b" ~ 0.6,
                           TRUE ~ value))

amod2 <- simFun_a(params = params2, 
       N0.a = N0.a, 
       N0.s = N0.s, 
       N0.p = N0.p, 
       L0 = L0, 
       simtime = years,
       alt.comp = T,
       alt.per = T,
       alt.seed = T)
```

```{r mod-2, tidy = F}
mod2 <- figfun(amod2, nmod, "./microstegium_model_kortessis_comparison_output_equal.pdf")
```

#### Use Amy's litter parameters (parameterization 3)
```{r amy-sim-3, tidy = F}
amod3 <- simFun_a(params = params, 
       N0.a = N0.a, 
       N0.s = N0.s, 
       N0.p = N0.p, 
       L0 = L0, 
       simtime = years,
       alt.comp = T,
       alt.per = T,
       alt.seed = T)
```

```{r mod-3, tidy = F}
mod3 <- figfun(amod3, nmod, "./microstegium_model_kortessis_comparison_output_aklitter.pdf")
```

#### Use Amy's competition parameterization (parameterization 4)
```{r amy-sim-4, tidy = F}
amod4 <- simFun_a(params = params2, 
       N0.a = N0.a, 
       N0.s = N0.s, 
       N0.p = N0.p, 
       L0 = L0, 
       simtime = years,
       alt.comp = F,
       alt.per = T,
       alt.seed = T)
```

```{r mod-4, tidy = F}
mod4 <- figfun(amod4, nmod, "./microstegium_model_kortessis_comparison_output_akcomp.pdf")
```

#### Use Amy's assumptions about first-year perennial seed production (parameterization 5)
```{r amy-sim-5, tidy = F}
amod5 <- simFun_a(params = params2, 
       N0.a = N0.a, 
       N0.s = N0.s, 
       N0.p = N0.p, 
       L0 = L0, 
       simtime = years,
       alt.comp = T,
       alt.per = F,
       alt.seed = T)
```

```{r mod-5, tidy = F}
mod5 <- figfun(amod5, nmod, "./microstegium_model_kortessis_comparison_output_akper.pdf")
```

#### Use Amy's assumption perennial survival before seed production (parameterization 6)
```{r amy-sim-6, tidy = F}
amod6 <- simFun_a(params = params2, 
       N0.a = N0.a, 
       N0.s = N0.s, 
       N0.p = N0.p, 
       L0 = L0, 
       simtime = years,
       alt.comp = T,
       alt.per = T,
       alt.seed = F)
```

```{r mod-6, tidy = F}
mod6 <- figfun(amod6, nmod, "./microstegium_model_kortessis_comparison_output_akseed.pdf")
```

