---
title: "microstegium_elymus_model_summary"
author: "Amy Kendig"
date: "12/5/2019"
output: 
  bookdown::pdf_document2: default
---

```{r options, include = F}
library(knitr)
# wrap text
opts_chunk$set(tidy.opts=list(width.cutoff = 70), tidy = T)
```

# Summary
Simulation of *Microstegium vimineum* and *Elymus virgincus* over time. *M. vimineum* is an invasive annual grass and *E. virginicus* is a native perennial grass. *E. virginicus* grows by itself for 100 years, and then *M. vimineum* is introduced. Disease effects arise 100 years later. Population dynamics are similar in the absence of disease and when both species are susceptible to disese. In both cases, *M. vimineum* reduces the size of the *E. virginicus* population. *M. vimineum* is able to recover population losses due to disease. *E. virginicus* seedlings are impacted more than adults. When only *M. vimineum* is susceptible to disease, its population crashes.


# Set-up
```{r set-up, message = F, warning = F}
# clear all existing data
rm(list=ls())

# open libraries
library(data.table)
library(plotly)
library(cowplot)
library(popbio)
library(tidyverse)

# plotting parameters
axisText=10
axisTitle=12
legendText=10
legendTitle=0
colPal = c("#0072B2", "#56B4E9", "#009E73", "#E69F00")

# import data
params <- read_csv("../data/mv_ev_model_parameters_120319.csv", col_types = cols())
```

# Parameters

## Formatting:
* subscripts follow "."
* A = annual plant
* L = annual litter
* P = perennial adult (at least 1 year old)
* S = perennial seedling (germinated that spring)  
* Composite parameters are capitalized  

## Settings:
```{r parameters, tidy = F}
# simulation time
simtime = 500

# invasion time
invtime = 100

# disease time
distime = invtime + 100

# initial conditions
N0.A = 0 # initial annual population size
N0.S = 1 # initial perennial seedling population size
N0.P = 0 # initial perennial adult population size
L0 = 0 # initial annual litter amount
Ni.A = 1 # introduction of annual 
```

## Literature:
```{r parameters-table, echo = F, results = 'asis'}
paramSub <- filter(params, !(symbol %in% c("alpha.AA", "alpha.AS", "alpha.AP", "alpha.SS", "alpha.SA", "alpha.SP", "alpha.PP", "alpha.PS", "alpha.PA")))
kable(paramSub, caption = "Model parameter values", format = "markdown", padding = 0)
```


# Model

## Population equations:
Assume counts are being conducted in the spring  
  
N.A[t+1] = s.A * (1-G.A) * N.A[t] + G.A * V.A * y.A * N.A[t]  
annual seeds = seed bank survival + seed production  

L[t+1] = a * G.A * V.A * N.A[t] + (1-b) * L[t]    
annual litter = biomass from previous fall + decomposition  
  
N.S[t+1] = s.S * (1-G.S) * N.S[t] + G.S * F.1 * N.S[t] + F.P * N.P[t]  
perennial seeds = seed bank survival + seedling seed production + adult seed production  
  
N.P[t+1] = s.P * N.P[t] + s.1 * G.S * N.S[t]  
perennial adults = survival + seedling maturation  

## Litter suppression of germination
G.A = g.A / (1 + beta * L[t])  
annual seed germination = germination in the absence of litter / reduction due to litter 

G.S = g.S / (1 + beta * L[t])  
perennial seed germination = germination in the absence of litter / reduction due to litter 

## Density-dependence on growth and fecundity
V.A = v.A / (1 + alpha.A * G.A * N.A[t] + alpha.P * N.P[t] + alpha.S * G.S * N.S[t])    
annual growth = growth in the absence of competition / (competition from annuals + competition from perennial adults + competition from perennial seedlings)

F.1 = f.1 / (1 + alpha.A * G.A * N.A[t] + alpha.P * N.P[t] + alpha.S * G.S * N.S[t])    
perennial seedling seed production = seed production in the absence of competition / (competition from annuals + competition from perennial adults + competition from perennial seedlings)

F.P = f.P / (1 + alpha.A * G.A * N.A[t] + alpha.P * N.P[t] + alpha.S * G.S * N.S[t])    
perennial adult seed production = seed production in the absence of competition / (competition from annuals + competition from perennial adults + competition from perennial seedlings)


# Function
```{r model}
simFun = function(params, N0.A, N0.S, N0.P, L0, Ni.A, simtime, invtime, distime){
  
  # define parameters
  g.A = filter(params, symbol == "g.A")$value
  beta = filter(params, symbol == "beta")$value
  s.A = filter(params, symbol == "s.A")$value
  v.A = filter(params, symbol == "v.A")$value
  y.A = filter(params, symbol == "y.A")$value
  a = filter(params, symbol == "a")$value
  b = filter(params, symbol == "b")$value
  g.S = filter(params, symbol == "g.S")$value
  s.S = filter(params, symbol == "s.S")$value
  f.P = filter(params, symbol == "f.P")$value
  f.1 = filter(params, symbol == "f.1")$value
  s.1 = filter(params, symbol == "s.1")$value
  s.P = filter(params, symbol == "s.P")$value
  alpha.A = filter(params, symbol == "alpha.A")$value
  alpha.S = filter(params, symbol == "alpha.S")$value
  alpha.P = filter(params, symbol == "alpha.P")$value
  dis = filter(params, symbol == "dis")$value

  # initialize populations
  N.A=rep(NA,simtime)
  N.S=rep(NA,simtime)
  N.P=rep(NA,simtime)
  L=rep(NA,simtime)
  
  N.A[1]=N0.A
  N.S[1]=N0.S
  N.P[1]=N0.P
  L[1]=L0
  
  # simulate population dynamics
  for(t in 1:(simtime - 1)){	
    
    # introduce annual at t=100
    N.A[t] = ifelse(N0.A == 0 & t == invtime, Ni.A, N.A[t])
    
    # calulate parameters to introduce disease at distime
    growth.A = ifelse(t < distime, v.A, dis*g.A)
    
    # reduce germination due to litter
    G.A = g.A / (1 + beta * L[t])
    G.S = g.S / (1 + beta * L[t])
    
    # reduce fecundity or growth due to competition
    V.A = growth.A / (1 + alpha.A * G.A * N.A[t] + alpha.P * N.P[t] + alpha.S * G.S * N.S[t])
    F.1 = f.1 / (1 + alpha.A * G.A * N.A[t] + alpha.P * N.P[t] + alpha.S * G.S * N.S[t])
    F.P = f.P / (1 + alpha.A * G.A * N.A[t] + alpha.P * N.P[t] + alpha.S * G.S * N.S[t])
    
    # population size
    N.A[t+1] = s.A * (1-G.A) * N.A[t] + G.A * V.A * y.A * N.A[t]
    L[t+1] = a * G.A * V.A * N.A[t] + (1-b) * L[t]
    N.S[t+1] = s.S * (1-G.S) * N.S[t] + G.S * F.1 * N.S[t] + F.P * N.P[t]  
    N.P[t+1] = s.P * N.P[t] + s.1 * G.S * N.S[t]  
    
    # correct to prevent negative numbers
    N.A[t+1] = ifelse(N.A[t+1] < 1, 0, N.A[t+1])
    L[t+1] = ifelse(L[t+1] < 1, 0, L[t+1])
    N.S[t+1] = ifelse(N.S[t+1] < 1, 0, N.S[t+1])
    N.P[t+1] = ifelse(N.P[t+1] < 1, 0, N.P[t+1])
  }
  
  # save data
  dfN = data.frame(time = rep(1:simtime, 4), N = c(N.S, N.P, N.A, L), species = rep(c("Elymus seedling", "Elymus adult", "Microstegium", "Microstegium litter"), each = simtime))
  
  # return
  return(dfN)
}
```

# Simulations

## No disease
```{r no-dis, tidy = F}
nodisSim <- simFun(params, N0.A, N0.S, N0.P, L0, Ni.A, simtime, invtime, distime = simtime + 10)
nodisSim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)
```

### Elymus on non-transformed scale
```{r elymus-zoom, tidy = F}
nodisSim %>%
  filter(species %in% c("Elymus adult", "Elymus seedling")) %>%
  ggplot(aes(x = time, y = N, color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)
```

## Add disease
```{r def-parms, tidy = F}
disSim <- simFun(params, N0.A, N0.S, N0.P, L0, Ni.A, simtime, invtime, distime)
disSim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)
```

### Microstegium on non-transformed scale
```{r micro-zoom, tidy = F}
disSim %>%
  filter(species %in% c("Microstegium", "Microstegium litter")) %>%
  ggplot(aes(x = time, y = N, color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal[3:4])
```

### Time to extinction?
```{r elymus-time, tidy = F}
nodisSim %>%
  filter(species == "Elymus seedling" & N == 0) %>%
  filter(time == min(time))
```

```{r micro-time, tidy = F}
disSim %>%
  filter(species == "Microstegium" & N == 0 & time > invtime) %>%
  filter(time == min(time))
```
