---
title: "microstegium_elymus_model_ideas"
author: "Amy Kendig"
date: "11/14/2019"
output: pdf_document
---

```{r options, include = F}
library(knitr)
# wrap text
opts_chunk$set(tidy.opts=list(width.cutoff = 70), tidy = T)
```

## Summary
Simulation of *Microstegium vimineum* and *Elymus virgincus* over time. *M. vimineum* is an invasive annual grass and *E. virginicus* is a native perennial grass. *E. virginicus* grows by itself for 100 years, and then *M. vimineum* is introduced. Disease effects arise 100 years later. Population dynamics are similar in the absence of disease and when both species are susceptible to disese. In both cases, *M. vimineum* reduces the size of the *E. virginicus* population. *M. vimineum* is able to recover population losses due to disease. *E. virginicus* seedlings are impacted more than adults. When only *M. vimineum* is susceptible to disease, its population crashes.

```{r set-up, message = F, warning = F, include = F}
# clear all existing data
rm(list=ls())

# open libraries
library(data.table)
library(plotly)
library(cowplot)
library(popbio)
library(tidyverse)

# plotting parameters
axisText=10
axisTitle=12
legendText=10
legendTitle=0

# import data
params <- read_csv("../data/mv_ev_model_parameters_102119.csv", col_types = cols())
aparams <- read_csv("../data/mv_ev_model_alt_parameters_101519.csv", col_types = cols())
```

## Parameters

* subscripts follow "."
* p = perennial adult (at least 1 year old)
* s = perennial seedling (germinated that spring)
* a = annual
* L = annual litter

```{r parameters, tidy = F, include = F}
# derive related parameters
dparams <- tibble(
  parameter = c("annual-adult competition", 
                "seedling-adult competition", 
                "adult-annual competition", 
                "adult-seedling competition", 
                "adult-adult competition", 
                "seedling seed production"),
  symbol = c("alpha.pa", 
             "alpha.ps", 
             "alpha.ap", 
             "alpha.sp", 
             "alpha.pp", 
             "lambda.s"),
  value = c(filter(params, symbol == "alpha.sa")$value/10, 
            filter(params, symbol == "alpha.sa")$value/10, 
            filter(params, symbol == "alpha.as")$value*10, 
            filter(params, symbol == "alpha.as")$value*10, 
            filter(params, symbol == "alpha.as")$value*10,
            filter(params, symbol == "lambda.p")$value/10)
) %>%
  mutate(units = c(rep("year^-1^", 5), "seeds year^-1^"),
         reference = "derived")

# merge with main parameters
params2 <- full_join(params, 
                     dparams, 
                     by = c("parameter", "symbol", "value", "units", "reference"))

# simulation time
years = 500
```

```{r parameters-table, echo = F, results = 'asis'}
kable(params2, caption = "Model parameter values")
kable(aparams, caption = "Alternative model parameter values")
```

## Model

#### Population equations:
Assume counts are being conducted in the fall  
  
N.s[t+1] = s.s * (1-g.s) * N.s[t] + g.s * h.s * f.s * N.s[t] + m.p * f.p * N.p[t]  
perennial seeds = seed bank survival + seedling seed production + adult seed production  
*seedling survival seems to be density-dependent (intraspecific facilitation, interspecific competition)*  
  
N.p[t+1] = m.p * N.p[t] + g.s * h. s * N.s[t]  
perennial adults = survival + seedling maturation  
  
N.a[t+1] = s.a * (1-g.a) * N.a[t] + g.a * h.a * f.a * N.a[t]  
annual seeds = seed bank survival + seed production  
*seed survival or germination seems to depend on infection*  
  
L[t+1] = c.a * g.a * h.a * N.a[t] + L[t] * e^-b^  
annual litter = biomass from previous fall + decomposition  
*c.a is probably a function of density*  

### Density-dependence on fecundity
f.s = lam.s / (1 + alpha.ss * g.s * h.s * N.s[t] + alpha.sp * m.p * N.p[t] + alpha.sa * g.a * h.a * N.a[t])  
perennial seedling fecundity = fecundity in the absence of competition / (perennial seedling competition + perennial adult competition + annual competition)

f.p = lam.p / (1 + alpha.ps * g.s * h.s * N.s[t] + alpha.pp * m.p * N.p[t] + alpha.pa * g.a * h.a * N.a[t])  
perennial adult fecundity = fecundity in the absence of competition / (perennial seedling competition + perennial adult competition + annual competition)

f.a = lam.a / (1 + alpha.as * g.s * h.s * N.s[t] + alpha.ap * m.p * N.p[t] + alpha.aa * g.a * h.a * N.a[t])  
annual fecundity = fecundity in the absence of competition / (perennial seedling competition + perennial adult competition + annual competition)  
*we don't currently have good estimates for competition coefficients*  
*density-dependence of fecundity changes with infection*  

#### Litter suppression
g.s = gamma.s + (alpha.sL * L[t])  
perennial seed germination = germination in the absence of litter + reduction due to litter (alpha.sL < 0, g.s constrained to >= 0)

g.a = gamma.a + (alpha.aL * L[t])  
annual seed germination = germination in the absence of litter + reduction due to litter (alpha.aL < 0, g.a constrained to >= 0)  
*litter production is correlated with more cleistogamous seeds in the first year*  