---
title: "microstegium_elymus_model_summary"
author: "Amy Kendig"
date: "07/24/2020"
output: 
  bookdown::pdf_document2: default
---

```{r options, include = F}
library(knitr)
# wrap text
opts_chunk$set(tidy.opts=list(width.cutoff = 70), tidy = T)
```

# Summary
Simulation of *Microstegium vimineum* and *Elymus virgincus* over time. *M. vimineum* is an invasive annual grass and *E. virginicus* is a native perennial grass, with two stages: seedling and adult. Simulations are conducted for each species alone and invading the other after it has had time to establish (100 years). Simulations are also conducted in which both species are introduced to the system at the same time. In the absence of disease, *M. vimineum* is predicted to persist alone, but it is out-competed by *E. virginicus*. In the presence of disease, the competitive hierarchy reverses: *E. virginicus* can persist alone, but it is outcompeted by *M. vimineum*. *Microstegium vimineum* experiences a net benefit from the presence of disease, likely due to lower intraspecific competition, even though its biomass production and survival are lower. *Elymus virgincus* also experiences losses in biomass and survival in the presence of disease, but intraspecific competition does not decrease. Results were consistent whether species invaded one another or were "co-introduced".

# Set-up
```{r set-up, message = F, warning = F}
# clear all existing data
rm(list=ls())

# open libraries
library(data.table)
library(plotly)
library(cowplot)
library(popbio)
library(tidyverse)

# plotting parameters
axisText=10
axisTitle=12
legendText=10
legendTitle=0
colPal = c("#0072B2", "#56B4E9", "#009E73", "#E69F00")

# import data
params <- read_csv("../data/mv_ev_model_parameters_experiment_072420.csv", col_types = cols())
params_dis <- read_csv("../data/mv_ev_model_disease_parameters_experiment_072420.csv", col_types = cols())
```

# Parameters

## Formatting
* subscripts follow "."
* A = annual plant
* L = annual litter
* P = perennial adult (at least 1 year old)
* S = perennial seedling (germinated that spring)  
* Composite parameters are capitalized  

## Settings
```{r parameters, tidy = F}
# simulation time
simtime = 10000

# invasion time
invtime = 6000

# initial conditions
N0.A = 1 # initial annual population size
N0.S = 1 # initial perennial seedling population size
N0.P = 0 # initial perennial adult population size
L0 = 0 # initial annual litter amount
Ni.A = 1 # introduction of annual 
Ni.S = 1 # introduction of perennial
```

## "No disease" parameters from experiments
```{r parameters-table, echo = F, results = 'asis'}
kable(params %>% select(-notes), caption = "Model parameter values without disease", format = "markdown", padding = 0)
```

## "Disease" parameters from experiments
```{r disparameters-table, echo = F, results = 'asis'}
kable(params_dis %>% select(-notes), caption = "Model parameter values with disease", format = "markdown", padding = 0)
```


# Model

## Population equations
Assume counts are being conducted in the spring  
  
N.A[t+1] = s.A * (1-G.A) * N.A[t] + G.A * h.A * Y.A * N.A[t]  
annual seeds = seed bank survival + seed production  

L[t+1] = a * G.A * h.A * V.A * N.A[t] + (1-b) * L[t]    
annual litter = biomass from previous fall + decomposition  
  
N.S[t+1] = s.S * (1-G.S) * N.S[t] + G.S * h.S * Y.S * N.S[t] + h.P * Y.P * N.P[t]  
perennial seeds = seed bank survival + seedling seed production + adult seed production  
  
N.P[t+1] = h.P * w.P * N.P[t] + G.S * h.S * w.S * N.S[t]  
perennial adults = survival + seedling maturation  

## Litter suppression of germination
G.A = g.A / (1 + beta.A * L[t])  
annual seed germination = germination in the absence of litter / reduction due to litter 

G.S = g.S / (1 + beta.S * L[t])  
perennial seed germination = germination in the absence of litter / reduction due to litter 

## Density-dependence on growth
V.A = v.A / (1 + alpha.AA * G.A * N.A[t] + alpha.AP * N.P[t] + alpha.AS * G.S * N.S[t])    
annual growth = growth in the absence of competition / (competition from annuals + competition from perennial adults + competition from perennial seedlings)

V.S = v.S / (1 + alpha.SA * G.A * N.A[t] + alpha.SP * N.P[t] + alpha.SS * G.S * N.S[t])    
perennial seedling growth = growth in the absence of competition / (competition from annuals + competition from perennial adults + competition from perennial seedlings)

V.P = v.P / (1 + alpha.PA * G.A * N.A[t] + alpha.PP * N.P[t] + alpha.PS * G.S * N.S[t])    
perennial adult growth = growth in the absence of competition / (competition from annuals + competition from perennial adults + competition from perennial seedlings)

## Seed production based on biomass
Y.A = V.A^y.A * y.Aint    
annual seed production = annual biomass with conversion factors (from a log-log regression)

Y.S = V.S^y.S * y.Sint - 1    
perennial seedling seed production = perennial seedling biomass with conversion factors (from a log-log regression, 1 added to seed numbers before log-transforming)

Y.P = V.P^y.P * y.Pint - 1    
perennial adult seed production = perennial adult biomass with conversion factors (from a log-log regression, 1 added to seed numbers before log-transforming)


# Function
```{r model}
simFun = function(params, N0.A, N0.S, N0.P, L0, Ni.A, Ni.S, simtime, invtime, invader){
  
  # define parameters
  g.A = filter(params, symbol == "g.A")$value
  beta.A = filter(params, symbol == "beta.A")$value
  s.A = filter(params, symbol == "s.A")$value
  h.A = filter(params, symbol == "h.A")$value
  v.A = filter(params, symbol == "v.A")$value
  alpha.AA = filter(params, symbol == "alpha.AA")$value
  alpha.AS = filter(params, symbol == "alpha.AS")$value
  alpha.AP = filter(params, symbol == "alpha.AP")$value
  y.A = filter(params, symbol == "y.A")$value
  y.Aint = filter(params, symbol == "y.Aint")$value
  a = filter(params, symbol == "a")$value
  b = filter(params, symbol == "b")$value
  g.S = filter(params, symbol == "g.S")$value
  beta.S = filter(params, symbol == "beta.A")$value
  s.S = filter(params, symbol == "s.S")$value
  v.S = filter(params, symbol == "v.S")$value
  v.P = filter(params, symbol == "v.P")$value
  alpha.SA = filter(params, symbol == "alpha.SA")$value
  alpha.SS = filter(params, symbol == "alpha.SS")$value
  alpha.SP = filter(params, symbol == "alpha.SP")$value
  alpha.PA = filter(params, symbol == "alpha.PA")$value
  alpha.PS = filter(params, symbol == "alpha.PS")$value
  alpha.PP = filter(params, symbol == "alpha.PP")$value
  y.S = filter(params, symbol == "y.S")$value
  y.Sint = filter(params, symbol == "y.Sint")$value
  y.P = filter(params, symbol == "y.P")$value
  y.Pint = filter(params, symbol == "y.Pint")$value
  h.S = filter(params, symbol == "h.S")$value
  h.P = filter(params, symbol == "h.P")$value
  w.S = filter(params, symbol == "w.S")$value
  w.P = filter(params, symbol == "w.P")$value

  # initialize populations
  N.A=rep(NA,simtime)
  N.S=rep(NA,simtime)
  N.P=rep(NA,simtime)
  L=rep(NA,simtime)
  
  N.A[1]=N0.A
  N.S[1]=N0.S
  N.P[1]=N0.P
  L[1]=L0
  
  # simulate population dynamics
  for(t in 1:(simtime - 1)){	
    
    # introduce invader
    N.A[t] = ifelse(t == invtime & invader == "annual", Ni.A, N.A[t])
    N.S[t] = ifelse(t == invtime & invader == "perennial", Ni.S, N.S[t])
    
    # reduce germination due to litter
    G.A = g.A / (1 + beta.A * L[t])
    G.S = g.S / (1 + beta.S * L[t])
  
    # reduce growth due to competition
    V.A = v.A / (1 + alpha.AA * G.A * N.A[t] + alpha.AP * N.P[t] + alpha.AS * G.S * N.S[t])    
    V.S = v.S / (1 + alpha.SA * G.A * N.A[t] + alpha.SP * N.P[t] + alpha.SS * G.S * N.S[t]) 
    V.P = v.P / (1 + alpha.PA * G.A * N.A[t] + alpha.PP * N.P[t] + alpha.PS * G.S * N.S[t])  
    
    # seed production based on biomass
    Y.A = V.A^y.A * y.Aint
    # Y.S = V.S^y.S * y.Sint 
    Y.S = V.S^y.S * y.Sint - 1
    Y.S = ifelse(Y.S < 0, 0, Y.S)
    # Y.P = V.P^y.P * y.Pint
    Y.P = V.P^y.P * y.Pint - 1
    Y.P = ifelse(Y.P < 0, 0, Y.P)
    
    # population size
    N.A[t+1] = s.A * (1-G.A) * N.A[t] + G.A * h.A * Y.A * N.A[t]  
    L[t+1] = a * G.A * h.A * V.A * N.A[t] + (1-b) * L[t]    
    N.S[t+1] = s.S * (1-G.S) * N.S[t] + G.S * h.S * Y.S * N.S[t] + h.P * Y.P * N.P[t]
    N.P[t+1] = h.P * w.P * N.P[t] + G.S * h.S * w.S * N.S[t]  
    
    # correct to prevent negative numbers
    N.A[t+1] = ifelse(N.A[t+1] < 1, 0, N.A[t+1])
    L[t+1] = ifelse(L[t+1] < 1, 0, L[t+1])
    N.S[t+1] = ifelse(N.S[t+1] < 1, 0, N.S[t+1])
    N.P[t+1] = ifelse(N.P[t+1] < 1, 0, N.P[t+1])
  }
  
  # save data
  dfN = data.frame(time = rep(1:simtime, 4), N = c(N.S, N.P, N.A, L), species = rep(c("Elymus seedling", "Elymus adult", "Microstegium", "Microstegium litter"), each = simtime))
  
  # return
  return(dfN)
}
```

# Simulations

## No disease, Microstegium by itself
```{r no-dis-mv-alone, tidy = F}
# run simulation
no_dis_mv_alone_sim <- simFun(params, N0.A, N0.S = 0, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "none")

# figure of time series
no_dis_mv_alone_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
no_dis_mv_alone_pop <- filter(no_dis_mv_alone_sim, species == "Microstegium" & time == max(time)) %>%
  mutate(disease = "Without disease",
         interaction = "alone")
```

## No disease, Microstegium invades Elymus
```{r no-dis-mv-invades, tidy = F}
# run simulation
no_dis_mv_invades_sim <- simFun(params, N0.A = 0, N0.S, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "annual")

# figure of time series
no_dis_mv_invades_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
no_dis_mv_invades_pop <- filter(no_dis_mv_invades_sim, species == "Microstegium" & time == max(time)) %>%
  mutate(disease = "Without disease",
         interaction = "invader")

no_dis_ev_resident_pop <- filter(no_dis_mv_invades_sim, species %in% c("Elymus adult", "Elymus seedling") & time == max(time)) %>%
  pivot_wider(names_from = species, values_from = N) %>%
  rename("Elymus_adult" = "Elymus adult", "Elymus_seedling" = "Elymus seedling") %>%
  mutate(species = "Elymus",
         N = Elymus_adult + Elymus_seedling,
         disease = "Without disease",
         interaction = "resident")
```

## No disease, co-introduced
```{r no-dis-co-intro, tidy = F}
# run simulation
no_dis_co_intro_sim <- simFun(params, N0.A = 100, N0.S = 100, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "none")

# figure of time series
no_dis_co_intro_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
no_dis_co_intro_mv_pop <- filter(no_dis_co_intro_sim, species == "Microstegium" & time == max(time)) %>%
  mutate(disease = "Without disease",
         interaction = "in competition")

no_dis_co_intro_ev_pop <- filter(no_dis_co_intro_sim, species %in% c("Elymus adult", "Elymus seedling") & time == max(time)) %>%
  pivot_wider(names_from = species, values_from = N) %>%
  rename("Elymus_adult" = "Elymus adult", "Elymus_seedling" = "Elymus seedling") %>%
  mutate(species = "Elymus",
         N = Elymus_adult + Elymus_seedling,
         disease = "Without disease",
         interaction = "in competition")
```

## Disease, Microstegium by itself
```{r dis-mv-alone, tidy = F}
# run simulation
dis_mv_alone_sim <- simFun(params_dis, N0.A, N0.S = 0, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "none")

# figure of time series
dis_mv_alone_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
dis_mv_alone_pop <- filter(dis_mv_alone_sim, species == "Microstegium" & time == max(time)) %>%
  mutate(disease = "With disease",
         interaction = "alone")
```

## Disease, Microstegium invades Elymus
```{r dis-mv-invades, tidy = F}
# run simulation
dis_mv_invades_sim <- simFun(params_dis, N0.A = 0, N0.S, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "annual")

# figure of time series
dis_mv_invades_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
dis_mv_invades_pop <- filter(dis_mv_invades_sim, species == "Microstegium" & time == max(time)) %>%
  mutate(disease = "With disease",
         interaction = "invader")

dis_ev_resident_pop <- filter(dis_mv_invades_sim, species %in% c("Elymus adult", "Elymus seedling") & time == max(time)) %>%
  pivot_wider(names_from = species, values_from = N) %>%
  rename("Elymus_adult" = "Elymus adult", "Elymus_seedling" = "Elymus seedling") %>%
  mutate(species = "Elymus",
         N = Elymus_adult + Elymus_seedling,
         disease = "With disease",
         interaction = "resident")
```

## Disease, co-introduced
```{r dis-co-intro, tidy = F}
# run simulation
dis_co_intro_sim <- simFun(params_dis, N0.A = 100, N0.S = 100, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "none")

# figure of time series
dis_co_intro_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
dis_co_intro_mv_pop <- filter(dis_co_intro_sim, species == "Microstegium" & time == max(time)) %>%
  mutate(disease = "With disease",
         interaction = "in competition")

dis_co_intro_ev_pop <- filter(dis_co_intro_sim, species %in% c("Elymus adult", "Elymus seedling") & time == max(time)) %>%
  pivot_wider(names_from = species, values_from = N) %>%
  rename("Elymus_adult" = "Elymus adult", "Elymus_seedling" = "Elymus seedling") %>%
  mutate(species = "Elymus",
         N = Elymus_adult + Elymus_seedling,
         disease = "With disease",
         interaction = "in competition")
```

## No disease, Elymus by itself
```{r no-dis-ev-alone, tidy = F}
# run simulation
no_dis_ev_alone_sim <- simFun(params, N0.A = 0, N0.S, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "none")

# figure of time series
no_dis_ev_alone_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
no_dis_ev_alone_pop <- filter(no_dis_ev_alone_sim, species %in% c("Elymus adult", "Elymus seedling") & time == max(time)) %>%
  pivot_wider(names_from = species, values_from = N) %>%
  rename("Elymus_adult" = "Elymus adult", "Elymus_seedling" = "Elymus seedling") %>%
  mutate(species = "Elymus",
         N = Elymus_adult + Elymus_seedling,
         disease = "Without disease",
         interaction = "alone")
```

## No disease, Elymus invades Microstegium
```{r no-dis-ev-invades, tidy = F}
# run simulation
no_dis_ev_invades_sim <- simFun(params, N0.A, N0.S = 0, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "perennial")

# figure of time series
no_dis_ev_invades_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
no_dis_ev_invades_pop <- filter(no_dis_ev_invades_sim, species %in% c("Elymus adult", "Elymus seedling") & time == max(time)) %>%
  pivot_wider(names_from = species, values_from = N) %>%
  rename("Elymus_adult" = "Elymus adult", "Elymus_seedling" = "Elymus seedling") %>%
  mutate(species = "Elymus",
         N = Elymus_adult + Elymus_seedling,
         disease = "Without disease",
         interaction = "invader")

no_dis_mv_resident_pop <- filter(no_dis_ev_invades_sim, species == "Microstegium" & time == max(time)) %>%
  mutate(disease = "Without disease",
         interaction = "resident")
```

## Disease, Elymus by itself
```{r dis-ev-alone, tidy = F}
# run simulation
dis_ev_alone_sim <- simFun(params_dis, N0.A = 0, N0.S, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "none")

# figure of time series
dis_ev_alone_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
dis_ev_alone_pop <- filter(dis_ev_alone_sim, species %in% c("Elymus adult", "Elymus seedling") & time == max(time)) %>%
  pivot_wider(names_from = species, values_from = N) %>%
  rename("Elymus_adult" = "Elymus adult", "Elymus_seedling" = "Elymus seedling") %>%
  mutate(species = "Elymus",
         N = Elymus_adult + Elymus_seedling,
         disease = "With disease",
         interaction = "alone")
```

## Disease, Elymus invades Microstegium
```{r dis-ev-invades, tidy = F}
# run simulation
dis_ev_invades_sim <- simFun(params_dis, N0.A, N0.S = 0, N0.P, L0, Ni.A, Ni.S, simtime, invtime, "perennial")

# figure of time series
dis_ev_invades_sim %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line(size = 1.5) + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle)) +
  scale_color_manual(values = colPal)

# save final density
dis_ev_invades_pop <- filter(dis_ev_invades_sim, species %in% c("Elymus adult", "Elymus seedling") & time == max(time)) %>%
  pivot_wider(names_from = species, values_from = N) %>%
  rename("Elymus_adult" = "Elymus adult", "Elymus_seedling" = "Elymus seedling") %>%
  mutate(species = "Elymus",
         N = Elymus_adult + Elymus_seedling,
         disease = "With disease",
         interaction = "invader")

dis_mv_resident_pop <- filter(dis_ev_invades_sim, species == "Microstegium" & time == max(time)) %>%
  mutate(disease = "With disease",
         interaction = "resident")
```

# Population Figure

```{r population-figure, tidy = F}

# combine data
pop_fig_dat <- no_dis_mv_alone_pop %>%
  full_join(no_dis_mv_invades_pop) %>%
  full_join(no_dis_ev_resident_pop) %>%
  full_join(dis_mv_alone_pop) %>%
  full_join(dis_mv_invades_pop) %>%
  full_join(dis_ev_resident_pop) %>%
  full_join(no_dis_ev_alone_pop) %>%
  full_join(no_dis_ev_invades_pop) %>%
  full_join(no_dis_mv_resident_pop) %>%
  full_join(dis_ev_alone_pop) %>%
  full_join(dis_ev_invades_pop) %>%
  full_join(dis_mv_resident_pop) %>%
  mutate(species = fct_relevel(species, "Microstegium"),
         N = N + 1,
         log_N = log(N),
         disease = fct_relevel(disease, "Without disease"))

# template theme
temp_theme <- theme_bw() +
  theme(axis.text = element_text(size = 10, color="black"),
        axis.title = element_text(size = 12),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.text = element_text(size = 10),
        legend.title = element_blank(),
        legend.box.margin = margin(-10, -10, -10, -10),
        legend.position = "bottom",
        legend.direction = "horizontal",
        strip.background = element_blank(),
        strip.text = element_text(size = 10, face = "italic"),
        strip.placement = "outside",
        plot.title = element_text(size = 12, hjust = 0.5))

# colors
col_pal = c("#55A48B", "#C0A76D")

# figure
pop_fig <- ggplot(pop_fig_dat, aes(x = interaction, y = N)) +
  geom_bar(stat = "identity", position = "dodge", aes(fill = disease)) +
  scale_fill_manual(values = col_pal) +
  facet_wrap(~species, nrow = 1, scales = "free") +
  xlab("Status") +
  ylab("Simulated long-term population size") +
  temp_theme
pop_fig

# save figure
pdf("../output/simulated_long_term_pop_figure.pdf", width = 6, height = 4)
pop_fig
dev.off()

```