---
title: "microstegium_elymus_model_summary"
author: "Amy Kendig"
date: "11/14/2019"
output: pdf_document
---

```{r options, include = F}
library(knitr)
# wrap text
opts_chunk$set(tidy.opts=list(width.cutoff = 70), tidy = T)
```

## Summary
Simulation of *Microstegium vimineum* and *Elymus virgincus* over time. *M. vimineum* is an invasive annual grass and *E. virginicus* is a native perennial grass. *E. virginicus* grows by itself for 100 years, and then *M. vimineum* is introduced. Disease effects arise 100 years later. Population dynamics are similar in the absence of disease and when both species are susceptible to disese. In both cases, *M. vimineum* reduces the size of the *E. virginicus* population. *M. vimineum* is able to recover population losses due to disease. *E. virginicus* seedlings are impacted more than adults. When only *M. vimineum* is susceptible to disease, its population crashes.


## Set-up
```{r set-up, message = F, warning = F}
# clear all existing data
rm(list=ls())

# open libraries
library(data.table)
library(plotly)
library(cowplot)
library(popbio)
library(tidyverse)

# plotting parameters
axisText=10
axisTitle=12
legendText=10
legendTitle=0

# import data
params <- read_csv("../data/mv_ev_model_parameters_102119.csv", col_types = cols())
aparams <- read_csv("../data/mv_ev_model_alt_parameters_101519.csv", col_types = cols())
```

## Parameters

#### Formatting:
* subscripts follow "."
* p = perennial adult (at least 1 year old)
* s = perennial seedling (germinated that spring)
* a = annual
* L = annual litter

#### Edit:
```{r parameters, tidy = F}
# derive related parameters
dparams <- tibble(
  parameter = c("annual-adult competition", 
                "seedling-adult competition", 
                "adult-annual competition", 
                "adult-seedling competition", 
                "adult-adult competition", 
                "seedling seed production"),
  symbol = c("alpha.pa", 
             "alpha.ps", 
             "alpha.ap", 
             "alpha.sp", 
             "alpha.pp", 
             "lambda.s"),
  value = c(filter(params, symbol == "alpha.sa")$value/10, 
            filter(params, symbol == "alpha.sa")$value/10, 
            filter(params, symbol == "alpha.as")$value*10, 
            filter(params, symbol == "alpha.as")$value*10, 
            filter(params, symbol == "alpha.as")$value*10,
            filter(params, symbol == "lambda.p")$value/10)
) %>%
  mutate(units = c(rep("year^-1^", 5), "seeds year^-1^"),
         reference = "derived")

# merge with main parameters
params2 <- full_join(params, 
                     dparams, 
                     by = c("parameter", "symbol", "value", "units", "reference"))

# simulation time
years = 500
```

```{r parameters-table, echo = F, results = 'asis'}
kable(params2, caption = "Model parameter values")
kable(aparams, caption = "Alternative model parameter values")
```

## Model

#### Population equations:
Assume counts are being conducted in the fall  
  
N.s[t+1] = s.s * (1-g.s) * N.s[t] + g.s * h.s * f.s * N.s[t] + m.p * f.p * N.p[t]  
perennial seeds = seed bank survival + seedling seed production + adult seed production  
  
N.p[t+1] = m.p * N.p[t] + g.s * h. s * N.s[t]  
perennial adults = survival + seedling maturation  
  
N.a[t+1] = s.a * (1-g.a) * N.a[t] + g.a * h.a * f.a * N.a[t]  
annual seeds = seed bank survival + seed production  
  
L[t+1] = c.a * g.a * h.a * N.a[t] + L[t] * e^-b^  
annual litter = biomass from previous fall + decomposition  

### Density-dependence on fecundity
f.s = lam.s / (1 + alpha.ss * g.s * h.s * N.s[t] + alpha.sp * m.p * N.p[t] + alpha.sa * g.a * h.a * N.a[t])  
perennial seedling fecundity = fecundity in the absence of competition / (perennial seedling competition + perennial adult competition + annual competition)

f.p = lam.p / (1 + alpha.ps * g.s * h.s * N.s[t] + alpha.pp * m.p * N.p[t] + alpha.pa * g.a * h.a * N.a[t])  
perennial adult fecundity = fecundity in the absence of competition / (perennial seedling competition + perennial adult competition + annual competition)

f.a = lam.a / (1 + alpha.as * g.s * h.s * N.s[t] + alpha.ap * m.p * N.p[t] + alpha.aa * g.a * h.a * N.a[t])  
annual fecundity = fecundity in the absence of competition / (perennial seedling competition + perennial adult competition + annual competition)

#### Litter suppression
g.s = gamma.s + (alpha.sL * L[t])  
perennial seed germination = germination in the absence of litter + reduction due to litter (alpha.sL < 0, g.s constrained to >= 0)

g.a = gamma.a + (alpha.aL * L[t])  
annual seed germination = germination in the absence of litter + reduction due to litter (alpha.aL < 0, g.a constrained to >= 0)

#### Function
```{r model}
simFun = function(params, N0.a, N0.s, N0.p, L0, Ni.a, simtime){
  
  # define parameters
  m.p=filter(params, symbol == "m.p")$value
  s.a=filter(params, symbol == "s.a")$value
  s.s=filter(params, symbol == "s.s")$value
  gamma.a=filter(params, symbol == "gamma.a")$value
  gamma.s=filter(params, symbol == "gamma.s")$value
  alpha.aL=filter(params, symbol == "alpha.aL")$value
  alpha.sL=filter(params, symbol == "alpha.sL")$value
  b=filter(params, symbol == "b")$value
  h.a=filter(params, symbol == "h.a")$value
  h.s=filter(params, symbol == "h.s")$value
  h.p=filter(params, symbol == "h.p")$value
  lambda.a=filter(params, symbol == "lambda.a")$value
  lambda.p=filter(params, symbol == "lambda.p")$value
  lambda.s=filter(params, symbol == "lambda.s")$value
  alpha.aa=filter(params, symbol == "alpha.aa")$value
  alpha.as=filter(params, symbol == "alpha.as")$value
  alpha.ap=filter(params, symbol == "alpha.ap")$value
  alpha.sa=filter(params, symbol == "alpha.sa")$value
  alpha.ss=filter(params, symbol == "alpha.ss")$value
  alpha.sp=filter(params, symbol == "alpha.sp")$value
  alpha.pa=filter(params, symbol == "alpha.pa")$value
  alpha.ps=filter(params, symbol == "alpha.ps")$value
  alpha.pp=filter(params, symbol == "alpha.pp")$value
  c.a=filter(params, symbol == "c.a")$value
  tol.a=filter(params, symbol == "tol.a")$value
  tol.p=filter(params, symbol == "tol.p")$value
  
  # initialize populations
  N.a=rep(NA,simtime)
  N.s=rep(NA,simtime)
  N.p=rep(NA,simtime)
  L=rep(NA,simtime)
  
  N.a[1]=N0.a
  N.s[1]=N0.s
  N.p[1]=N0.p
  L[1]=L0
  
  # simulate population dynamics
  for(t in 1:(simtime-1)){	
    
    # introduce annual at t=100
    N.a[t]=ifelse(N0.a==0 & t==100, Ni.a, N.a[t])
    
    # calulate parameters to introduce disease at t=200
    lam.a=ifelse(t<200, lambda.a, tol.a*lambda.a)
    lam.p=ifelse(t<200, lambda.p, tol.p*lambda.p)
    lam.s=ifelse(t<200, lambda.s, tol.p*lambda.s)
    
    # reduce germination due to litter
    g.s=gamma.s+(alpha.sL*L[t])
    g.s=ifelse(g.s<0, 0, g.s)
    g.a=gamma.a+(alpha.aL*L[t])
    g.a=ifelse(g.a<0, 0, g.a)
    
    # reduce fecundity due to competition
    f.s=lam.s/(1+alpha.ss*g.s*h.s*N.s[t]+alpha.sp*m.p*N.p[t]+alpha.sa*g.a*h.a*N.a[t])
    f.p=lam.p/(1+alpha.ps*g.s*h.s*N.s[t]+alpha.pp*m.p*N.p[t]+alpha.pa*g.a*h.a*N.a[t])
    f.a=lam.a/(1+alpha.as*g.s*h.s*N.s[t]+alpha.ap*m.p*N.p[t]+alpha.aa*g.a*h.a*N.a[t])
    
    # population size
    N.s[t+1]=s.s*(1-g.s)*N.s[t]+g.s*h.s*f.s*N.s[t]+m.p*f.p*N.p[t]
    N.p[t+1]=m.p*N.p[t]+g.s*h.s*N.s[t]
    N.a[t+1]=s.a*(1-g.a)*N.a[t]+g.a*h.a*f.a*N.a[t]
    L[t+1]=c.a*g.a*h.a*N.a[t]+L[t]*exp(-b)
    
    # correct to prevent negative numbers
    N.s[t+1]=ifelse(N.s[t+1]<1,0,N.s[t+1])
    N.p[t+1]=ifelse(N.p[t+1]<1,0,N.p[t+1])
    N.a[t+1]=ifelse(N.a[t+1]<1,0,N.a[t+1])
    L[t+1]=ifelse(L[t+1]<0,0,L[t+1])
  }
  
  # save data
  dfN=data.frame(time=rep(1:simtime,4),N=c(N.s,N.p,N.a,L),species=rep(c("Elymus seedling","Elymus adult","Microstegium","Microstegium litter"),each=simtime))
  
  # return
  return(dfN)
}
```

## Simulations

#### Default parameters
```{r def-parms, tidy = F}
simFun(params = params2, 
       N0.a = 0, 
       N0.s = 1, 
       N0.p = 1, 
       L0 = 0, 
       Ni.a = 1, 
       simtime=years) %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line() + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle))
```

### Remove disease from both species
```{r rem-dis, tidy = F}
simFun(params = params2 %>% mutate(value = case_when(symbol %in% c("tol.a", "tol.p") ~ 1,
                                                     TRUE ~ value)),
       N0.a = 0, 
       N0.s = 1, 
       N0.p = 1, 
       L0 = 0, 
       Ni.a = 1, 
       simtime=years) %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line() + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle))
```

### Remove disease from perennial
```{r rem-dis-per, tidy = F}
simFun(params = params2 %>% mutate(value = case_when(symbol == "tol.p" ~ 1,
                                                     TRUE ~ value)),
       N0.a = 0, 
       N0.s = 1, 
       N0.p = 1, 
       L0 = 0, 
       Ni.a = 1, 
       simtime=years) %>%
  ggplot(aes(x = time, y = log(N+1), color = species)) +
  geom_line() + 
  theme(axis.text = element_text(size = axisText),
        axis.title = element_text(size = axisTitle),
        legend.text = element_text(size = legendText),
        legend.title = element_text(size = legendTitle))
```